Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-08-27 manuscript.tex
pushing final version of manuscript

\hypertarget{toc}{}
\tableofcontents
\newpage
\section[Introduction]{Precise temporal localisation of M/EEG effects
with Bayesian generalised additive multilevel models}

@@

The issues with multiple comparisons represent a common and
well-recognised danger in neuroimaging and M/EEG research, where the
collected data allows for a multitude of potential hypothesis tests and
is characterised by complex structures of spatiotemporal dependencies.
The probability of obtaining at least one false positive in an ensemble
(family) of \(m\) tests (i.e., the FWER) is computed as
\(1-\left(1-\alpha\right)^{m}\) (for \(m=10\) independent tests and
\(\alpha=0.05\), it is approximately equal to \(0.4\)). Different
methods exist to control the FWER, that is, to bring it back to
\(\alpha\). Most methods apply a simple correction to series of
\(p\)-values issued from univariate statistical tests (e.g., t-tests).
For instance, the Bonferroni correction (\citeproc{ref-dunn1961}{Dunn,
1961}) consists in setting the significance threshold to \(\alpha/m\),
or equivalently, multiplying the \(p\)-values by \(m\) and using the
standard \(\alpha\) significance threshold. This method is generally
overconservative (i.e., under-powered) as it assumes statistical
independence of the tests, an assumption that is clearly violated in the
context of M/EEG timeseries characterised by massive spatiotemporal
dependencies. Some alternative methods aims at controlling the FDR,
defined as the proportion of false positive \emph{among positive tests}
(e.g., \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995};
\citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}). However, a
major limitation of both types of corrections is that they do not take
into account the spatial and temporal information contained in M/EEG
data.

@@

We then fitted a Bayesian GAM (BGAM) using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018})
and default priors (i.e., weakly informative priors). We ran eight
Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\) (\citeproc{ref-gabry2019}{Gabry et al., 2019};
\citeproc{ref-gelman2020}{Gelman et al., 2020}). The \texttt{brms}
package uses the same syntax as the \texttt{R} package \texttt{mgcv} v
1.9-3 (\citeproc{ref-mgcv}{Wood, 2017b}) for specifying smooth effects.
Figure~\ref{fig-plot-post-slope} shows the predictions of this model
together with the raw data.

However, the previous model only included constant (fixed) effects, thus
not properly accounting for between-participant variability. We next
fitted a multilevel version of the BGAM (BGAMM, for an introduction to
Bayesian multilevel models in \texttt{brms}, see
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}) including a
varying intercept and slope for participant (but with a constant
smoother). Although it is possible to fit a BGAMM using data at the
single-trial level, we present a computationally lighter version of the
model that is fitted directly on by-participant summary statistics (mean
and SD), similar to what is done in meta-analysis.

@@

\fontsize{9.0pt}{10.8pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lcccc}
\toprule
 & Bias & MAE & RMSE & Variance \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{onset} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
BGAM (our method) & 0.11 & 2.76 & 0.11 & 28.78 \\ 
Change point & 5.17 & 6.51 & 5.17 & 33.58 \\ 
Cluster mass & 6.64 & 7.62 & 6.64 & 207.86 \\ 
Holm & 29.83 & 32.01 & 29.83 & 2,763.68 \\ 
TFCE & 29.73 & 32.07 & 29.73 & 2,823.11 \\ 
FDR BY01 & 36.67 & 50.80 & 36.67 & 7,872.30 \\ 
FDR BH95 & 58.64 & 102.49 & 58.64 & 19,054.58 \\ 
Raw p-value & 70.72 & 132.86 & 70.72 & 25,004.65 \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{offset} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
BGAM (our method) & -2.35 & 3.46 & 2.35 & 53.63 \\ 
Cluster mass & -8.64 & 9.44 & 8.64 & 193.63 \\ 
Change point & -9.28 & 9.82 & 9.28 & 33.03 \\ 
Holm & -31.86 & 34.03 & 31.86 & 2,764.12 \\ 
TFCE & -31.84 & 34.19 & 31.84 & 2,834.11 \\ 
FDR BY01 & -36.68 & 51.37 & 36.68 & 7,648.53 \\ 
FDR BH95 & -58.87 & 102.63 & 58.87 & 18,939.58 \\ 
Raw p-value & -72.93 & 133.33 & 72.93 & 25,012.72 \\ 
\bottomrule
\end{tabular*}

@@

In brief, our results show that the model-based approach we introduced
outperforms conventional methods such as cluster-based inference in
identifying the onset and offset of M/EEG effects. Performance was
assessed both on simulated data--allowing us to evaluate the method's
ability to recover ground-truth onset and offset values--and on actual
MEG data--allowing us to assess its sensitivity to realistic data
properties (subset similarity). Together, these results highlight
desirable properties for any method aiming to precisely and reliably
estimate the onset and offset of M/EEG effects: it should i) recover
true onsets and offsets in simulation (good asymptotic behaviour), ii)
identify clusters that are interpretable and consistent in empirical
data, and iii) show sensitivity to subtle changes in the data. Our
approach meets all three of these desiderata.

@@

\textbf{I am not sure what to do with this paragraph (not sure how to
discuss these results/observations):} Interestingly, the TFCE method
performed worse than the traditional cluster-sum approach, consistent
with the predictions of Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}) based on the original
findings of S. Smith \& Nichols (\citeproc{ref-smith2009}{2009}). We
also found a striking overlap in the clusters identified by the
\texttt{Holm} and \texttt{TFCE} procedures (cf.
Figure~\ref{fig-onset-offset} and Figure~\ref{fig-splits-similarity}).
Whereas these two approaches are conceptually distinct; \texttt{Holm}
controlling the family-wise error rate through sequential \emph{p}-value
adjustment, \texttt{TFCE} enhancing signal based on spatiotemporal
support; their similarity in practice may arise because both effectively
prioritise extended, moderately strong effects over isolated
high-intensity points. This convergence warrants further methodological
work. We chose not to include the cluster-depth algorithm
(\citeproc{ref-frossard2022}{Frossard \& Renaud, 2022}), as prior work
already shows its performance to be inferior to cluster-mass approaches
in similar contexts (\citeproc{ref-rousselet_using_2025}{Rousselet,
2025}).

A critical consideration for any model-based approach is that the model
must adequately capture the underlying data-generating process.
Misspecified models are likely to produce biased or unreliable
onset/offset estimates. This underscores the importance of thorough
model diagnostics, including posterior predictive checks, fit
assessments, and model comparison (\citeproc{ref-gelman2020}{Gelman et
al., 2020}). As with any inferential modelling framework, transparency
in assumptions and rigorous validation procedures are indispensable to
ensure replicability and generalisability of the results.

Another important methodological consideration concerns the selection of
model hyperparameters, such as the threshold for posterior odds and the
number of basis functions. Although our simulations suggest that these
parameters influence the precision and reliability of onset and offset
estimates, optimal values may vary depending on the signal's temporal
dynamics and signal-to-noise characteristics. Future work could explore
principled approaches to hyperparameter tuning, including
cross-validation or fully Bayesian model selection using tools such as
leave-one-out cross-validation (LOO-CV) or Bayes factors
(\citeproc{ref-gelman2020}{Gelman et al., 2020}). We provide initial
guidance in \hyperref[apx-basis]{Appendix~B} and advocate for future
development of adaptive heuristics to support flexible yet parsimonious
model specification.

@@

We used R version 4.4.3 (\citeproc{ref-base}{R Core Team, 2025}) and the
following R packages: assertthat v. 0.2.1
(\citeproc{ref-assertthat}{Wickham, 2019}), brms v. 2.22.0
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018},
\citeproc{ref-brms2021}{2021}), doParallel v. 1.0.17
(\citeproc{ref-doParallel}{Corporation \& Weston, 2022}), easystats v.
0.7.4 (\citeproc{ref-easystats}{Lüdecke et al., 2022}), foreach v. 1.5.2
(\citeproc{ref-foreach}{Microsoft \& Weston, 2022}), furrr v. 0.3.1
(\citeproc{ref-furrr}{Vaughan \& Dancho, 2022}), future v. 1.58.0
(\citeproc{ref-future}{Bengtsson, 2021}), ggpubr v. 0.6.0
(\citeproc{ref-ggpubr}{Kassambara, 2023}), ggrepel v. 0.9.6
(\citeproc{ref-ggrepel}{Slowikowski, 2024}), glue v. 1.8.0
(\citeproc{ref-glue}{Hester \& Bryan, 2024}), grateful v. 0.2.12
(\citeproc{ref-grateful}{Rodriguez-Sanchez \& Jackson, 2024}), gt v.
1.0.0 (\citeproc{ref-gt}{Iannone et al., 2025}), knitr v. 1.50
(\citeproc{ref-knitr2014}{Xie, 2014}, \citeproc{ref-knitr2015}{2015},
\citeproc{ref-knitr2025}{2025}), MetBrewer v. 0.2.0
(\citeproc{ref-MetBrewer}{Mills, 2022}), mgcv v. 1.9.3
(\citeproc{ref-mgcv2003}{Wood, 2003b}, \citeproc{ref-mgcv2004}{2004},
\citeproc{ref-mgcv2011}{2011}, \citeproc{ref-mgcv2017}{2017c};
\citeproc{ref-mgcv2016}{Wood et al., 2016}), neurogam v. 0.0.1
(\citeproc{ref-neurogam}{Nalborczyk, 2025}), pakret v. 0.2.2
(\citeproc{ref-pakret}{Gallou, 2024}), patchwork v. 1.3.0
(\citeproc{ref-patchwork}{T. L. Pedersen, 2024}), rmarkdown v. 2.29
(\citeproc{ref-rmarkdown2024}{Allaire et al., 2024};
\citeproc{ref-rmarkdown2018}{Xie et al., 2018},
\citeproc{ref-rmarkdown2020}{2020}), scales v. 1.4.0
(\citeproc{ref-scales}{Wickham et al., 2025}), scico v. 1.5.0
(\citeproc{ref-scico}{T. L. Pedersen \& Crameri, 2023}), signal v. 1.8.1
(\citeproc{ref-signal}{signal developers, 2023}), tictoc v. 1.2.1
(\citeproc{ref-tictoc}{Izrailev, 2024}), tidybayes v. 3.0.7
(\citeproc{ref-tidybayes}{Kay, 2024}), tidytext v. 0.4.2
(\citeproc{ref-tidytext}{Silge \& Robinson, 2016}), tidyverse v. 2.0.0
(\citeproc{ref-tidyverse}{Wickham et al., 2019}).

@@

\bibitem[\citeproctext]{ref-easystats}
Lüdecke, D., Ben-Shachar, M. S., Patil, I., Wiernik, B. M., Bacher, E.,
Thériault, R., \& Makowski, D. (2022). {easystats}: Framework for easy
statistical modeling, visualization, and reporting. \emph{CRAN}.
\url{https://doi.org/10.32614/CRAN.package.easystats}

@@

@@

To model cross-temporal generalisation matrices of decoding performance
(ROC AUC), we extended our initial BGAM to take into account the
bivariate temporal distribution of AUC values, thus producing naturally
smoothed estimates (timecourses) of AUC values and posterior
probabilities. This model can be written as follows:

\[
\begin{aligned}
\text{AUC}_{i} &\sim \mathrm{Beta}(\mu_{i}, \phi)\\
g(\mu_{i}) &= f \left(\text{train}_{i}, \text{test}_{i} \right)\\
\end{aligned}
\]

@@

where we assume that AUC values come from a \(\mathrm{Beta}\)
distribution with two parameters \(\mu\) and \(\phi\). We can think of
\(f \left(\text{train}_{i}, \text{test}_{i} \right)\) as a surface (a
smooth function of two variables) that we can model using a
2-dimensional splines. Let
\(\mathbf{s}_{i} = \left(\text{train}_{i}, \text{test}_{i} \right)\) be
some pair of training and testing samples, and let
\(\mathbf{k}_{m} = \left(\text{train}_{m}, \text{test}_{m} \right)\)
denote the \(m^{\text{th}}\) knot in the domain of \(\text{train}_{i}\)
and \(\text{test}_{i}\). We can then express the smooth function as:

\[
f \left(\text{train}_{i}, \text{test}_{i} \right) = \alpha + \sum_{m=1}^M \beta_{m} b_{m} \left(\tilde{s}_{i}, \tilde{k}_{m} \right)
\]

@@

A popular bivariate basis function uses \emph{thin-plate splines}
(\citeproc{ref-wood2003}{Wood, 2003a}). These splines are designed to
interpolate and approximate smooth surfaces over two dimensions (hence
the ``bivariate'' term). For \(d=2\) dimensions and \(l=2\) (smoothness
penalty involving second order derivative):

\[
f \left(\tilde{s}_{i} \right) = \alpha + \beta_{1} x_{i} + \beta_{2} z_{i} +\sum_{m=1}^{M} \beta_{2+m} b_m\left(\tilde{s}_i, \tilde{k}_m\right)
\]

using the the radial basis function given by:

\[
b_m\left(\tilde{s}_i, \tilde{k}_m\right)=\left\|\tilde{s}_i-\tilde{k}_m\right\|^2 \log \left\|\tilde{s}_i-\tilde{k}_m\right\|
\]

where \(\left\|\mathbf{s}_i-\mathbf{k}_{m}\right\|\) is the Euclidean
distance between the covariate \(\mathbf{s}_{i}\) and the knot location
\(\mathbf{k}_{m}\). We fitted this model using \texttt{brms} and the
\texttt{t2()} tensor product smooth constructor with full penalties
(\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al.,
2019}). We ran eight MCMCs to approximate the posterior distribution,
including each 5000 iterations and a warmup of 2000 iterations, yielding
a total of \(8 \times (5000-2000) = 24000\) posterior samples to be used
for inference.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting a GAM with two temporal dimensions}
\NormalTok{timegen\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# 2D thin{-}plate spline (tp) with full penalties}
\NormalTok{    auc }\SpecialCharTok{\textasciitilde{}} \FunctionTok{t2}\NormalTok{(train\_time, test\_time, }\AttributeTok{bs =} \StringTok{"tp"}\NormalTok{, }\AttributeTok{k =} \DecValTok{30}\NormalTok{, }\AttributeTok{full =} \ConstantTok{TRUE}\NormalTok{),}
    \AttributeTok{data =}\NormalTok{ timegen\_data,}
    \AttributeTok{family =}\NormalTok{ brms}\SpecialCharTok{::}\FunctionTok{Beta}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.95}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{15}\NormalTok{),}
    \AttributeTok{backend =} \StringTok{"cmdstanr"}\NormalTok{,}
    \AttributeTok{stan\_model\_args =} \FunctionTok{list}\NormalTok{(}\AttributeTok{stanc\_options =} \FunctionTok{list}\NormalTok{(}\StringTok{"O1"}\NormalTok{) ),}
    \AttributeTok{file =} \StringTok{"models/timegen\_gam\_meg\_t2\_k30\_full.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

Figure~\ref{fig-gam-timegen-post-preds} shows the predictions from the
model (left) superimposed with the identified cluster as defined by
thresholding the posterior odds (right). Notably, this model could be
extended to a multilevel bivariate GAM via
\texttt{t2(train\_time,\ test\_time,\ participant,\ bs\ =\ c("tp",\ "tp",\ "re"),\ m\ =\ 2,\ full\ =\ TRUE)}
and could be generalised to account for both spatial (\texttt{x1} and
\texttt{x2}) and temporal (\texttt{time}) dimensions with formulas such
as \texttt{te(x,\ y,\ time,\ d\ =\ c(2,\ 1)\ )}. The performance of such
models should be assessed in future work.

@@

>>>>

@@

The issues with multiple comparisons represent a common and
well-recognised danger in neuroimaging and M/EEG research, where the
collected data allows for a multitude of potential hypothesis tests and
is characterised by complex structures of spatiotemporal dependencies.
The probability of obtaining at least one false positive in an ensemble
(family) of \(m\) independent tests (i.e., the FWER) is computed as
\(1-\left(1-\alpha\right)^{m}\) (for \(m=10\) independent tests and
\(\alpha=0.05\), it is approximately equal to \(0.4\)). Different
methods exist to control the FWER, that is, to bring it back to
\(\alpha\). Most methods apply a simple correction to series of
\(p\)-values issued from univariate statistical tests (e.g., t-tests).
For instance, the Bonferroni correction (\citeproc{ref-dunn1961}{Dunn,
1961}) consists in setting the significance threshold to \(\alpha/m\),
or equivalently, multiplying the \(p\)-values by \(m\) and using the
standard \(\alpha\) significance threshold. This method is generally
overconservative (i.e., under-powered) as it assumes statistical
independence of the tests, an assumption that is clearly violated in the
context of M/EEG timeseries characterised by massive spatiotemporal
dependencies. Some alternative methods aims at controlling the FDR,
defined as the proportion of false positive \emph{among positive tests}
(e.g., \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995};
\citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}). However, a
major limitation of both types of corrections is that they do not take
into account the spatial and temporal information contained in M/EEG
data.

@@

We then fitted a Bayesian GAM (BGAM) using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018})
and default priors (i.e., weakly informative priors). We ran eight
Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\) (\citeproc{ref-gabry2019}{Gabry et al., 2019};
\citeproc{ref-gelman2020}{Gelman et al., 2020};
\citeproc{ref-vehtari2021}{Vehtari et al., 2021}). The \texttt{brms}
package uses the same syntax as the \texttt{R} package \texttt{mgcv} v
1.9-3 (\citeproc{ref-mgcv}{Wood, 2017b}) for specifying smooth effects.
Figure~\ref{fig-plot-post-slope} shows the predictions of this model
together with the raw data.

However, the model whose predictions are depicted in
Figure~\ref{fig-plot-post-slope} only included constant (fixed) effects,
thus not properly accounting for between-participant variability. We
next fitted a multilevel version of the BGAM (BGAMM, for an introduction
to Bayesian multilevel models in \texttt{brms}, see
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}) including a
varying intercept and slope for participant (but with a constant
smoother). Although it is possible to fit a BGAMM using data at the
single-trial level, we present a computationally lighter version of the
model that is fitted directly on by-participant summary statistics (mean
and SD), similar to what is done in meta-analysis.

@@

\fontsize{9.0pt}{10.8pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lcccc}
\toprule
 & Bias & MAE & RMSE & Variance \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{onset} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
BGAM (our method) & {\bfseries 0.11} & {\bfseries 2.76} & {\bfseries 0.11} & {\bfseries 28.78} \\ 
Change point & 5.17 & 6.51 & 5.17 & 33.58 \\ 
Cluster mass & 6.64 & 7.62 & 6.64 & 207.86 \\ 
Holm & 29.83 & 32.01 & 29.83 & 2,763.68 \\ 
TFCE & 29.73 & 32.07 & 29.73 & 2,823.11 \\ 
FDR BY01 & 36.67 & 50.80 & 36.67 & 7,872.30 \\ 
FDR BH95 & 58.64 & 102.49 & 58.64 & 19,054.58 \\ 
Raw p-value & 70.72 & 132.86 & 70.72 & 25,004.65 \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{offset} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
BGAM (our method) & {\bfseries -2.35} & {\bfseries 3.46} & {\bfseries 2.35} & 53.63 \\ 
Cluster mass & -8.64 & 9.44 & 8.64 & 193.63 \\ 
Change point & -9.28 & 9.82 & 9.28 & {\bfseries 33.03} \\ 
Holm & -31.86 & 34.03 & 31.86 & 2,764.12 \\ 
TFCE & -31.84 & 34.19 & 31.84 & 2,834.11 \\ 
FDR BY01 & -36.68 & 51.37 & 36.68 & 7,648.53 \\ 
FDR BH95 & -58.87 & 102.63 & 58.87 & 18,939.58 \\ 
Raw p-value & -72.93 & 133.33 & 72.93 & 25,012.72 \\ 
\bottomrule
\end{tabular*}

@@

In brief, our results show that the model-based approach we introduced
outperforms conventional cluster-based methods in identifying the onset
and offset of M/EEG effects. We first assessed the performance of this
approach on simulated data allowing us to evaluate the method's ability
to recover ground-truth onset and offset values. We then assessed its
performance on actual MEG data, allowing us to assess its sensitivity to
realistic data properties (subset similarity). Together, these results
highlight desirable properties for any method aiming to precisely and
reliably estimate the onset and offset of M/EEG effects: it should i)
recover true onsets and offsets in simulation (good asymptotic
behaviour), ii) identify clusters that are interpretable and consistent
in empirical data, and iii) show sensitivity to subtle changes in the
data. Our approach meets all three of these desiderata.

@@

Interestingly, the TFCE method performed worse than the traditional
cluster-sum approach, consistent with the predictions of Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}) based on the original
findings of S. Smith \& Nichols (\citeproc{ref-smith2009}{2009}). We
also found a striking overlap in the clusters identified by the
\texttt{Holm} and \texttt{TFCE} procedures (cf.
Figure~\ref{fig-onset-offset} and Figure~\ref{fig-splits-similarity}).
Whereas these two approaches are conceptually distinct; \texttt{Holm}
controlling the family-wise error rate through sequential \emph{p}-value
adjustment, \texttt{TFCE} enhancing signal based on spatiotemporal
support; their similarity in practice may arise because both effectively
prioritise extended, moderately strong effects over isolated
high-intensity points. This convergence warrants further methodological
work.

A critical requirement for any model-based approach is that the model
must adequately capture the underlying data-generating process.
Misspecified models are likely to produce biased or unreliable
onset/offset estimates. This underscores the importance of thorough
model diagnostics, including posterior predictive checks, fit
assessments, and model comparison (\citeproc{ref-gelman2020}{Gelman et
al., 2020}). An important and related methodological consideration
concerns the selection of model hyperparameters, such as the number of
basis functions and the threshold for posterior odds. Although our
simulations suggest that these parameters influence the precision and
reliability of onset and offset estimates, optimal values may vary
depending on the signal's temporal dynamics and signal-to-noise
characteristics. Future work could explore principled approaches to
hyperparameter tuning, including cross-validation or fully Bayesian
model selection using tools such as leave-one-out cross-validation
(LOO-CV) or Bayes factors (\citeproc{ref-gelman2020}{Gelman et al.,
2020}). We provide initial guidance in \hyperref[apx-basis]{Appendix~B}
and advocate for future development of adaptive heuristics to support
flexible yet parsimonious model specification.

@@

We used R version 4.4.3 (\citeproc{ref-base}{R Core Team, 2025}) and the
following R packages: assertthat v. 0.2.1
(\citeproc{ref-assertthat}{Wickham, 2019}), brms v. 2.22.0
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018},
\citeproc{ref-brms2021}{2021}), doParallel v. 1.0.17
(\citeproc{ref-doParallel}{Corporation \& Weston, 2022}), foreach v.
1.5.2 (\citeproc{ref-foreach}{Microsoft \& Weston, 2022}), furrr v.
0.3.1 (\citeproc{ref-furrr}{Vaughan \& Dancho, 2022}), future v. 1.58.0
(\citeproc{ref-future}{Bengtsson, 2021}), ggpubr v. 0.6.0
(\citeproc{ref-ggpubr}{Kassambara, 2023}), ggrepel v. 0.9.6
(\citeproc{ref-ggrepel}{Slowikowski, 2024}), glue v. 1.8.0
(\citeproc{ref-glue}{Hester \& Bryan, 2024}), grateful v. 0.2.12
(\citeproc{ref-grateful}{Rodriguez-Sanchez \& Jackson, 2024}), gt v.
1.0.0 (\citeproc{ref-gt}{Iannone et al., 2025}), knitr v. 1.50
(\citeproc{ref-knitr2014}{Xie, 2014}, \citeproc{ref-knitr2015}{2015},
\citeproc{ref-knitr2025}{2025}), MetBrewer v. 0.2.0
(\citeproc{ref-MetBrewer}{Mills, 2022}), mgcv v. 1.9.3
(\citeproc{ref-mgcv2003}{Wood, 2003b}, \citeproc{ref-mgcv2004}{2004},
\citeproc{ref-mgcv2011}{2011}, \citeproc{ref-mgcv2017}{2017c};
\citeproc{ref-mgcv2016}{Wood et al., 2016}), neurogam v. 0.0.1
(\citeproc{ref-neurogam}{Nalborczyk, 2025}), pakret v. 0.2.2
(\citeproc{ref-pakret}{Gallou, 2024}), patchwork v. 1.3.0
(\citeproc{ref-patchwork}{T. L. Pedersen, 2024}), rmarkdown v. 2.29
(\citeproc{ref-rmarkdown2024}{Allaire et al., 2024};
\citeproc{ref-rmarkdown2018}{Xie et al., 2018},
\citeproc{ref-rmarkdown2020}{2020}), scales v. 1.4.0
(\citeproc{ref-scales}{Wickham et al., 2025}), scico v. 1.5.0
(\citeproc{ref-scico}{T. L. Pedersen \& Crameri, 2023}), signal v. 1.8.1
(\citeproc{ref-signal}{signal developers, 2023}), tictoc v. 1.2.1
(\citeproc{ref-tictoc}{Izrailev, 2024}), tidybayes v. 3.0.7
(\citeproc{ref-tidybayes}{Kay, 2024}), tidytext v. 0.4.2
(\citeproc{ref-tidytext}{Silge \& Robinson, 2016}), tidyverse v. 2.0.0
(\citeproc{ref-tidyverse}{Wickham et al., 2019}).

@@

@@

\bibitem[\citeproctext]{ref-vehtari2021}
Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., \& Bürkner, P.-C.
(2021). Rank-normalization, folding, and localization: An improved Rˆ
for assessing convergence of MCMC (with discussion). \emph{Bayesian
Analysis}, \emph{16}(2). \url{https://doi.org/10.1214/20-ba1221}

@@

To model cross-temporal generalisation matrices of decoding performance,
we extended our initial BGAM to take into account the bivariate temporal
distribution of AUC values, thus producing naturally smoothed estimates
(timecourses) of AUC values and posterior odds. This model can be
written as follows:

\[
\begin{aligned}
\text{AUC}_{i} &\sim \mathrm{Beta} \left(\mu_{i}, \phi\right)\\
\operatorname{logit} \left(\mu_{i} \right) &= \alpha + f \left(\text{train}_{i}, \text{test}_{i} \right)\\
f \left(\text{train}_{i}, \text{test}_{i} \right) &\equiv \sum_{k=1}^K b_{k} B_{k}\left(\text{train}_{i}, \text{test}_{i} \right)
\end{aligned}
\]

@@

where AUC values are assumed to follow a \(\mathrm{Beta}\) distribution,
parametrised by a mean \(\mu_{i}\) and a precision parameter \(\phi\).
The mean \(\mu_{i}\) is linked to the predictors through a logit link
function. The smooth function \(f(\text{train}_{i}, \text{test}_{i})\)
represents a two-dimensional surface defined over training and testing
times, which captures how decoding performance varies across the
temporal generalisation matrix. This surface is approximated by a linear
combination of \(K\) basis functions \(B_{k}(\cdot, \cdot)\), each
weighted by a coefficient \(b_{k}\). The basis functions are constructed
using a tensor product of univariate splines (here thin-plate splines)
applied to the training and testing time dimensions
(\citeproc{ref-wood2003}{Wood, 2003a}, \citeproc{ref-wood2017}{2017a}).

@@

We fitted this model using \texttt{brms} and the \texttt{t2()} tensor
product smooth constructor with full penalties
(\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al., 2019};
\citeproc{ref-wood2017}{Wood, 2017a}). We ran eight MCMCs to approximate
the posterior distribution, including each 5000 iterations and a warmup
of 1000 iterations, yielding a total of \(8 \times (5000-1000) = 32000\)
posterior samples to be used for inference.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting a GAM with two temporal dimensions}
\NormalTok{timegen\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# 2D thin{-}plate spline (tp) with full penalties}
\NormalTok{    auc }\SpecialCharTok{\textasciitilde{}} \FunctionTok{t2}\NormalTok{(train\_time, test\_time, }\AttributeTok{bs =} \StringTok{"tp"}\NormalTok{, }\AttributeTok{k =} \DecValTok{30}\NormalTok{, }\AttributeTok{full =} \ConstantTok{TRUE}\NormalTok{),}
    \AttributeTok{data =}\NormalTok{ timegen\_data,}
    \AttributeTok{family =} \FunctionTok{Beta}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.95}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{15}\NormalTok{)}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

Figure~\ref{fig-gam-timegen-post-preds} shows the predictions from the
model (left) superimposed with the identified cluster as defined by
thresholding the posterior odds (right). Notably, this model could be
extended to a multilevel bivariate GAM via
\texttt{t2(train\_time,\ test\_time,\ participant,\ bs\ =\ c("tp",\ "tp",\ "re"),\ m\ =\ 2,\ full\ =\ TRUE)}
and could be generalised to account for both spatial (\texttt{x} and
\texttt{y}) and temporal (\texttt{time}) dimensions with formulas such
as \texttt{te(x,\ y,\ time,\ d\ =\ c(2,\ 1)\ )}.

@@

\clearpage
\thispagestyle{empty}
\null
